The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
rewrote pruning section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2546 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 53a831f790efcf703d5b8d9663cad912d7e00d42

%\newpage
\subsection{Pruning}
\label{sec:pruning}
In early developed methods of unbiased recursive partitioning such as 
FACT \citep{Loh+Vanichsetakul:1988}, QUEST \citep{Loh+Shih:1997}, 
CRUISE \citep{Kim+Loh:2001} or GUIDE \citep{Loh:2002}
the $p$-values of applied significance tests are used purely for the selection of a split variable 
based on a unified scale. All other steps are performed in the same way as in classic tree 
algorithms, such as finding the best split point by optimizing an overall goodness-of-fit measure.

In subsequent approaches as in \cite{Hothorn+Hornik+Zeileis:2006}
it is shown that these $p$-values can also be used as a basis for a
pruning strategy namely \textit{prepruning}: Only if the $p$-value corresponding to the selected split
variable (hence the smallest $p$-value) is smaller than a given significance level a split is
performed while otherwise no further splits are made in the current node.
In that way the information provided by the significance test is exploited in a more profound way.
However, prepruning requires the applied test to perform reasonably well as a significance test, 
i.e. it has to have power but also avoid overfitting.

On the contrary, when employing postpruning no significance level is included leading to very large 
trees which are then pruned afterwards, for example by applying cost-complexity pruning as in the 
CART algorithm \citep{Breiman+Friedman+Stone:1984}. 
Thus, for postpruning the test only needs to be able to select the best split variable but does 
not have to be effective as a significance test.

Originally the GUIDE algorithm was introduced applying postpruning while CTree and MOB
were both presented employing prepruning (see Table~\ref{tab:combinations} for a more detailed
overview). Alternatively, for model-based partitioning a pruning strategy can also use model-based 
information criteria such as AIC or BIC \citep[see for example][]{Su+Wang+Fan:2004}.
However, all testing strategies can be combined with each of these pruning strategies.
If a test performs well the choice of a pruning strategy has hardly any impact on the results, 
but as prepruning only requires already provided information it is usually less computationally
expensive than postpruning needing additional calculations. 
However, if the test performs poorly in terms of power and overfitting, prepruning suffers from 
this deficiency while postpruning can compensate it at least partly and improve the results.
Therefore, being aware of the power of the selected testing strategy is essential for choosing 
an appropriate pruning strategy.
For that reason we are focusing on the testing strategies applied in tree algorithms 
and investigate how effective they are as significance tests.

>>>>

%\newpage
\subsection{Pruning}
\label{sec:pruning}

To avoid overfitting recursive partitioning algorithms need to assure that trees do not grow
too large. Apart from certain minimal subgroup size or maximal tree depth constraints this is
classically accomplished by so-called ``pruning'' approaches. As the three tree algorithms
considered here (CTree, MOB, and GUIDE) differ in their default choice of pruning approach,
we briefly discuss these here. However, as all three algorithms can in principle be combined
with any of the pruning approaches this is done only relatively briefly.

The classical CART algorithm \citep{Breiman+Friedman+Stone:1984} proposed to first grow a large
tree and then prune those splits in the tree that did not increase predictive performance in a
cross-validation. This is also know as \emph{postpruning} (after growing the initial tree) and
more specifically cost-complexity pruning.

In the unbiased recursive partitioning literature this postpruning approach is also usef frequently
\citep[e.g., in][]{Loh+Vanichsetakul:1988,Loh+Shih:1997,Kim+Loh:2001,Loh:2002} and the $p$-values
from the association tests are only employed for selecting the split variable on a unified scale.
However, \cite{Hothorn+Hornik+Zeileis:2006} proposed to also use these $p$-values for a 
so-called \emph{prepruning} strategy which stops growing the tree as soon as no significant association
can be found in a given subgroup. This approach is the default in CTree and also in
MOB. However, \cite{Zeileis+Hothorn+Hornik:2008} also pointed out that a natural strategy for
postpruning in model-based partitioning is to use information criteria such as AIC (Akaike information
criterion) or BIC (Bayes information criterion), following the ideas of \cite{Su+Wang+Fan:2004}.

Clearly, for inference-based prepruning it is crucial that the association tests employed for
split selection work well as statistical significance test, i.e., conform with their nominal size
and have high power. In contrast, when using postpruning (either based on cross-validation or
information criteria) it might not be as crucial that the significance test works well and has
high power.

Due to these considerations we first evaluate the significance tests underlying CTree, MOB,
and GUIDE by themselves, i.e., without growing an actual tree in combination with a pruning strategy.
Subsequently we combine the tests with a cost-complexity postpruning approach in order to assess
whether shortcomings of the tests are mitigated by pruning.
